---
title: "Hw1"
author: "Nivedita Attada"
date: "2024-06-11"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



## Part 1
1. What is an R object?
An R object holds data that can be a variable or information about variables.

2. How many ways can data be assigned to an R object?
"<-", "=", "assign()"

3. Why do you think it is important to ensure that data objects are of the correct type?
If objects are not the correct type, we may not be able to perform certain functions or get object mismatch errors when we do try to perform operations.

4. What is the relationship between vectors, matrices, and data frames?
Vectors are simple data types and contain one type of data, matrices are 2-dimensional and contain one data type, data frames can contain different data types in the same object.

5. Why might a data frame be more suitable than a matrix for holding heterogeneous biological data?
Biological data may contain different data types of data for example cell types and cell count would be two different data types and data frames are the only ones that can hold both types of variables

## Part 2

```{r part2}
values <- c(89, 90, 78, 81, 62, 
            78, 99, 94, 83, 51, 
            77, 85, 99, 80, 99, 
            56, 97, 87, 79, 88)

# Create the matrix
transcriptome <- matrix(values, nrow = 5, byrow = TRUE)

# Assign column names
colnames(transcriptome) <- c("Control", "Nitrogen", "Phosphate", "Potassium")

# Assign row names
rownames(transcriptome) <- c("GeneA", "GeneB", "GeneC", "GeneD", "GeneE")

# Print the matrix
print(transcriptome)

# Calculate the average expression for each gene
expression_average <- rowMeans(transcriptome)

# Print the average expression vector
print(expression_average)

# Order the rows of the matrix by the average expression values in descending order
sorted_indices <- order(expression_average, decreasing = TRUE)

# Reorder the matrix
sorted_transcriptome <- transcriptome[sorted_indices, ]

# Convert the sorted matrix into a data frame
sorted_genes <- as.data.frame(sorted_transcriptome)

# Print the sorted data frame
print(sorted_genes)
```

## Part 2

```{r part3, echo=TRUE}
# Set the working directory to the location of the file
#setwd("nattada/NYU/Transcriptomics/Hw1_Nivedita/")

# Load the file into a data frame
expvalues_df <- read.table("expvalues.txt", header = TRUE, sep = " ")

# Print the loaded data frame
head(expvalues_df)

# Assuming the first three columns are control and the last three are treatment
control_columns <- expvalues_df[, 1:3]
treatment_columns <- expvalues_df[, 4:6]

# Calculate the mean for control and treatment groups
control_means <- apply(control_columns, 1, mean)
treatment_means <- apply(treatment_columns, 1, mean)

# Add new columns control and treatment means
expvalues_df$Control_Mean <- control_means
expvalues_df$Treatment_Mean <- treatment_means

# Print the mean values
head(expvalues_df)

# Calculate fold change (treatment mean / control mean)
fold_change <- treatment_means / control_means

# Append the fold change to the data frame
expvalues_df$Fold_Change <- fold_change

# Print the mean values with fold change
head(expvalues_df)

# Calculate log2 fold change
log2_fold_change <- log2(fold_change)

# Append the log2 fold change to the data frame
expvalues_df$LFC <- log2_fold_change

# Print the mean values with fold change
head(expvalues_df)

# Count genes with LFC > 1 or LFC < -1
genes_lfc_gt_1 <- sum(expvalues_df$LFC > 1)
genes_lfc_lt_minus1 <- sum(expvalues_df$LFC < -1)

# Total genes with LFC > 1 or LFC < -1
genes_with_lfc <- genes_lfc_gt_1 + genes_lfc_lt_minus1

# Print the results
cat("Genes with Log2 Fold Change > 1:", genes_lfc_gt_1, "\n")
cat("Genes with Log2 Fold Change < -1:", genes_lfc_lt_minus1, "\n")
cat("Total genes with Log2 Fold Change > 1 or < -1:", genes_with_lfc, "\n")

# Extract genes with LFC > 1
induced_genes <- rownames(expvalues_df)[expvalues_df$LFC > 1]

# Write column names to file
writeLines(induced_genes, "Induced_genes.txt")

# Create a subset of data for induced genes only
induced_data <- expvalues_df[induced_genes, 1:6]

head(induced_data)

# Prepare PDF device for plotting
pdf("boxplot.pdf")

# Create boxplot
boxplot(induced_data, 
        main = "Expression Distribution of Induced Genes",
        xlab = "Experiments",
        ylab = "Expression Level",
        col = rainbow(ncol(induced_data)))

# Close the PDF device
dev.off()
```

